### 2018-04-26 contents/reviews/dmri.tex
Updates from ShareLaTeX

Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain white matter fibers \textit{in-vivo}, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques.

@@

DTI is based on the Gaussian model of water diffusion, and therefore imposes an important limitation. The Gaussian tensor can derive only one diffusion direction at a given voxel, and therefore is incapable of modelling cross-fiber diffusion of two or more directions. By extension, its derived metrics (FA, AD, RD) can only characterize a single diffusion direction as well. Therefore the interpretation of these metrics in white matter that contains cross-fiber configurations is tenuous.

@@

An end-to-end spherical deconvolution diffusion model was first proposed by Tournier et al. \cite{Tournier2004}. 
Consider a fiber population with distinct fiber components. The measured diffusion signal $S(\theta,\phi)$ can be written as:

\begin{equation}
S(\theta, \phi) = \sum_{i}^{} f_i \hat{A}_i R(\theta)
\end{equation}
where $\theta$ is the elevation angle,  $\phi$ the azimuth angle in spherical coordinates; $f_i$ is the volume fraction of the $i$-th fiber component, $\hat{A}$ is the rotation operator that orients fiber, and $R(\theta)$ is an axially symmetric response function.

$S(\theta,\phi) $ can be rewritten as a convolution over the unit sphere of $R(\theta)$ with a fiber orientation density function (FOD) $F(\theta, \phi)$, such that:
\begin{equation}
S(\theta, \phi) = F(\theta, \phi) \otimes R(\theta)
\end{equation}

With this approach, the FOD can be estimated using the reverse spherical deconvolution, using spherical and rotational harmonics \cite{Healy1998}. Spherical harmonics are orthonormal basis functions over a sphere, and therefore can approximate any symmetric diffusion geometry in the spherical q-space. The rotational harmonics similarly form the orthonormal basis over all possible rotations, and can represent any symmetric fiber configurations. The harmonics have two important parameters: the harmonic order $l$ and phase factor $m$, with $ -l \leq m \leq l $ for each harmonic order level $l$. The odd $l$ harmonics represent antisymmetric functions, while the even $l$ harmonics represent symmetric functions (\ref{fig:harmonics}).

@@

To reduce the number of parameters to be computed, we can make the further assumption that the diffusion process is symmetric, and therefore the odd $l$ harmonic orders can be ignored, with only the even harmonics considered. 
The response function $R(\theta)$ can be directly estimated from the dataset by supplying DWI region with high anisotropy from a single fiber bundle, such as the body of the corpus callosum.

@@

The family of methods that propagates by deterministic algorithms based on local orientation data is collectively called deterministic tractography. The first tracking algorithms were applied to DTI using step-wise polyline propagators that followed local DT tensors \cite{Conturo1999}. Further refinement of the line propagator to include neighbouring DT tensors, and interpolation of the local DT directions in order to obtain a smooth curve \cite{Mori2002b}. Basser introduced Runge-Kutta integration to allow more robust path-tracing when faced with noise in the local directions \cite{Basser2000}. Another similar path-integral method was also proposed by Tuch \cite{Tuch2000d}. The integrated tractography approaches are now standard practice for deterministic tractography \cite{Tournier2004,Pieper2004}. Deterministic tractography lends itself naturally to anatomical interpretation and further geometric analysis. It was traditionally paired with DTI, and so was thought to be limited in the case of crossing fibers.

Other methods attempted to circumvent the limitation with probabilistic approaches that include white matter tissue priors in the fiber propagation model. The most popular of the probabilistic approach is the FSL software's probabilistic tractography module (FDT) \cite{Behrens2007}, which also includes a model-based ball-and-stick diffusion model that can resolve 2-fiber crossings. The FDT software produces visitation images of its probabilistic streams that collate the probability of tract propagation at each voxel. The visitation image, however, is hard to interpret and localize, and therefore does not rival deterministic methods in surgical visualization.

The availability of higher order HARDI methods, especially in CSD allows the reinterpretation of FOD as local fiber orientation probability distribution \cite{Tournier2010,Jeurissen2011b}. Development of a probabilistic streamline propagator based on CSD can now combine the advantage of probabilistic models with the geometric interpretability of deterministic tractography.

>>>>

Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain white matter fibres \textit{in-vivo}, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques.

@@

DTI is based on the Gaussian model of water diffusion, and therefore imposes an important limitation. The Gaussian tensor can derive only one diffusion direction at a given voxel, and therefore is incapable of modelling cross-fibre diffusion of two or more directions. By extension, its derived metrics (FA, AD, RD) can only characterize a single diffusion direction as well. Therefore the interpretation of these metrics in white matter that contains cross-fibre configurations is tenuous.

@@

An end-to-end spherical deconvolution diffusion model was first proposed by Tournier et al. \cite{Tournier2004}. 
Consider a fibre population with distinct fibre components. The measured diffusion signal $S(\theta,\phi)$ can be written as:

\begin{equation}
S(\theta, \phi) = \sum_{i}^{} f_i \hat{A}_i R(\theta)
\end{equation}
where $\theta$ is the elevation angle,  $\phi$ the azimuth angle in spherical coordinates; $f_i$ is the volume fraction of the $i$-th fibre component, $\hat{A}$ is the rotation operator that orients fibre, and $R(\theta)$ is an axially symmetric response function.

$S(\theta,\phi) $ can be rewritten as a convolution over the unit sphere of $R(\theta)$ with a fibre orientation density function (FOD) $F(\theta, \phi)$, such that:
\begin{equation}
S(\theta, \phi) = F(\theta, \phi) \otimes R(\theta)
\end{equation}

With this approach, the FOD can be estimated using the reverse spherical deconvolution, using spherical and rotational harmonics \cite{Healy1998}. Spherical harmonics are orthonormal basis functions over a sphere, and therefore can approximate any symmetric diffusion geometry in the spherical q-space. The rotational harmonics similarly form the orthonormal basis over all possible rotations, and can represent any symmetric fibre configurations. The harmonics have two important parameters: the harmonic order $l$ and phase factor $m$, with $ -l \leq m \leq l $ for each harmonic order level $l$. The odd $l$ harmonics represent antisymmetric functions, while the even $l$ harmonics represent symmetric functions (\ref{fig:harmonics}).

@@

To reduce the number of parameters to be computed, we can make the further assumption that the diffusion process is symmetric, and therefore the odd $l$ harmonic orders can be ignored, with only the even harmonics considered. 
The response function $R(\theta)$ can be directly estimated from the dataset by supplying DWI region with high anisotropy from a single fibre bundle, such as the body of the corpus callosum.

@@

The family of methods that propagates by deterministic algorithms based on local orientation data is collectively called deterministic tractography. The first tracking algorithms were applied to DTI using step-wise polyline propagators that followed local DT tensors \cite{Conturo1999}. Further refinement of the line propagator to include neighbouring DT tensors, and interpolation of the local DT directions in order to obtain a smooth curve \cite{Mori2002b}. Basser introduced Runge-Kutta integration to allow more robust path-tracing when faced with noise in the local directions \cite{Basser2000}. Another similar path-integral method was also proposed by Tuch \cite{Tuch2000d}. The integrated tractography approaches are now standard practice for deterministic tractography \cite{Tournier2004,Pieper2004}. Deterministic tractography lends itself naturally to anatomical interpretation and further geometric analysis. It was traditionally paired with DTI, and so was thought to be limited in the case of crossing fibres.

Other methods attempted to circumvent the limitation with probabilistic approaches that include white matter tissue priors in the fibre propagation model. The most popular of the probabilistic approach is the FSL software's probabilistic tractography module (FDT) \cite{Behrens2007}, which also includes a model-based ball-and-stick diffusion model that can resolve 2-fibre crossings. The FDT software produces visitation images of its probabilistic streams that collate the probability of tract propagation at each voxel. The visitation image, however, is hard to interpret and localize, and therefore does not rival deterministic methods in surgical visualization.

The availability of higher order HARDI methods, especially in CSD allows the reinterpretation of FOD as local fibre orientation probability distribution \cite{Tournier2010,Jeurissen2011b}. Development of a probabilistic streamline propagator based on CSD can now combine the advantage of probabilistic models with the geometric interpretability of deterministic tractography.

### 2018-04-26 contents/reviews/pain.tex
Updates from ShareLaTeX

The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells in the spinal dorsal horn that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

@@

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}. The regions have been shown to consistently decrease in activity during attention-demanding, and non-self-referential tasks,when compared with activities during relaxed non-task states \cite{Raichle2015}. The DMN is believed to instantiates processes that support self-referential activities, emotional processing, and recollection \cite{Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), dorsal medial prefrontal cortex (dMPFC), posterior cingulate, inferior parietal, lateral temporal cortex, and hippocampal formation. Regions such as vmPFC and dMPFC overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. Pain related DMN deactivation have been found to be attenuated  in mind wandering away from pain, and may reflect antinociceptive processes \cite{Kucyi2013}, and that the DMN may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

The insula is an important neuroanatomical structure that has been implicated in both acute \cite{Lorenz2005,Starr2009} and chronic pain \cite{May2008,Jensen2016}. Anatomically it is subdivided into anterior and posterior components by the central insular sulcus. It can be further identified by its gyral features, consisting of 5 major gyri: anterior, middle and posterior short gyri in the anterior insula, and anterior long, posterior long gyri of the posterior insula. These gyri are not found consistently in the general population, but rather show individual variability \cite{Rosen2015}. Craig postulates that the insula is a key region for homeostasis and interoception \cite{Craig2003b}, and is necessary for pain discriminative sensations via VMpo projections in the middle insula in humans \cite{Craig2003a}. The anteior insula has been shown by functional connectivity analysis to have interactions with anteior MCC and middle insula with posterior MCC \cite{Taylor2009}. Further suggest that the anterior insula and MCC are involved in an emotional/salience network for interoception and emotional integration and monitoring \cite{Vogt2016}.

@@

Neuropathic pain can also arise from changes in the central neuroaxis. Trauma and other disruptions to the spinal pathways can affect noxious signal transmission. Where repeated and sustained noxious stimuli may cause the central nervous system to adapt and amplify responses to stimulation. These can involve activation-dependence plasticity in neurons of the dorsal horn, which result in windup of action potential discharge of these neurons. Additionally, long-term depression may also result in the inhibitory mechanism in interneurons of the substanita gelatinosa \cite{Woolf2000}. White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis. It is believed that injury to white matter fibers can affect action potential conduction between cortical and subcortical structures. This may give rise or contribute to neuropathic pain \cite{Geha2008}.

>>>>

The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells in the spinal dorsal horn that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibres inhibit, while small fibres facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

@@

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}. The regions have been shown to consistently decrease in activity during attention-demanding, and non-self-referential tasks,when compared with activities during relaxed non-task states \cite{Raichle2015}. The DMN is believed to instantiate processes that support self-referential activities, emotional processing, and recollection \cite{Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), dorsal medial prefrontal cortex (dMPFC), posterior cingulate, inferior parietal, lateral temporal cortex, and hippocampal formation. Regions such as vmPFC and dMPFC overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. Pain related DMN deactivation have been found to be attenuated  in mind wandering away from pain, and may reflect antinociceptive processes \cite{Kucyi2013}, and that the DMN may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

The insula is an important neuroanatomical structure that has been implicated in both acute \cite{Lorenz2005,Starr2009} and chronic pain \cite{May2008,Jensen2016}. Anatomically it is subdivided into anterior and posterior components by the central insular sulcus. It can be further identified by its gyral features, consisting of 5 major gyri: anterior, middle and posterior short gyri in the anterior insula, and anterior long, posterior long gyri of the posterior insula. These gyri are not found consistently in the general population, but rather show individual variability \cite{Rosen2015}. Craig postulates that the insula is a key region for homeostasis and interoception \cite{Craig2003b}, and is necessary for pain discriminative sensations via VMpo projections in the middle insula in humans \cite{Craig2003a}. The anterior insula has been shown by functional connectivity analysis to have interactions with anterior MCC and middle insula with posterior MCC \cite{Taylor2009}. Further suggest that the anterior insula and MCC are involved in an emotional/salience network for interoception and emotional integration and monitoring \cite{Vogt2016}.

@@

Neuropathic pain can also arise from changes in the central neuroaxis. Trauma and other disruptions to the spinal pathways can affect noxious signal transmission. Where repeated and sustained noxious stimuli may cause the central nervous system to adapt and amplify responses to stimulation. These can involve activation-dependence plasticity in neurons of the dorsal horn, which result in windup of action potential discharge of these neurons. Additionally, long-term depression may also result in the inhibitory mechanism in interneurons of the substantia gelatinosa \cite{Woolf2000}. White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis. It is believed that injury to white matter fibres can affect action potential conduction between cortical and subcortical structures. This may give rise or contribute to neuropathic pain \cite{Geha2008}.

### 2018-04-26 contents/reviews/trigeminal.tex
Updates from ShareLaTeX

The trigeminal nerve, or the fifth cranial nerve (CN V) is the primary afferent for both facial somatosensation and pain (\ref{fig:trig-branches}). It consists of, from superior to inferior, the ophthalmic (V1), maxillary (V2) and mandibular (V3) branches. The distribution of the branches consists of both motor and sensory components. The motor branch of CN V activates the muscles responsible for mastication, as well as tensor tympani, tensor velipalatini, mylohyoid and anterior belly of the digastric.

@@

Trigeminal sensory afferents include the myelinated A and unmyelinated C-fibers that originate from the subcutaneous sensory receptors responsible for mechano- and temperature pain, fine touch and vibration, such as the Messiner's, Pacinian and Ruffini corpuscles.

@@

CN V passes into the pons (Figure \ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem. It then curve towards the medial aspect, and synapses onto the trigeminal nucleus in the basis of the pontine brainstem. The trigeminal brainstem sensory nucleus complex is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei. The spinal nucleus can be further divided into pars oralis, pars interpolaris and pars caudalis.

@@

The main sensory nucleus receives discriminative input from the orofacial region, and along with the spinal trigeminal nucleus fibers, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract. A separate branch responsible for oral cavity, however, projects to the ipsilateral VC via the dorsal trigeminothalamic tract.

@@

Neuroimaging of the TN has been challenging, due to the small diameter of the CN V and since the contrast of the CNS trigeminal pathways is difficult to distinguish from the surrounding on conventional MRI \cite{Borges2010}. In conventional MRI, the brainstem nerve segment can only be deduced from its surrounding structures \cite{Casselman2008}. Sequences that can visualize myelinated structures are chosen, these include T2-weighted MRI  sequences such as FIESTA and CISS. Manual segmentation of T1 and T2 MRI requires great effort, with mixed results \cite{Miller2008a}. Diffusion MRI provided fiber orientation, and early attempts to segment the trigeminal system showed promise \cite{Upadhyay2008,Habas2007f}. Additionally, DTI showed strong ability to visualize spatial relationships in white matter anatomy, for example in delineating fornix sub-regions \cite{Chen2015c}, and in visualizing the distortion of CN VII/VIII by vestibular schwannomas \cite{Chen2011b}.

>>>>

The trigeminal nerve, or the fifth cranial nerve (CN V) is the primary afferent for both facial somatosensation and pain (Figure \ref{fig:trig-branches}). It consists of, from superior to inferior, the ophthalmic (V1), maxillary (V2) and mandibular (V3) branches. The distribution of the branches consists of both motor and sensory components. The motor branch of CN V activates the muscles responsible for mastication, as well as tensor tympani, tensor velipalatini, mylohyoid and anterior belly of the digastric.

@@

Trigeminal sensory afferents include the myelinated A and unmyelinated C-fibres that originate from the subcutaneous sensory receptors responsible for mechano- and temperature pain, fine touch and vibration, such as the Messiner's, Pacinian and Ruffini corpuscles.

@@

CN V passes into the pons (Figure \ref{fig:trig-pathway}), through the transverse pontine fibres and enters the tegmentum of the pontine brainstem. It then curve towards the medial aspect, and synapses onto the trigeminal nucleus in the basis of the pontine brainstem. The trigeminal brainstem sensory nucleus complex is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei. The spinal nucleus can be further divided into pars oralis, pars interpolaris and pars caudalis.

@@

The main sensory nucleus receives discriminative input from the orofacial region, and along with the spinal trigeminal nucleus fibres, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract. A separate branch responsible for oral cavity, however, projects to the ipsilateral VC via the dorsal trigeminothalamic tract.

@@

Neuroimaging of the TN has been challenging, due to the small diameter of the CN V and since the contrast of the CNS trigeminal pathways is difficult to distinguish from the surrounding on conventional MRI \cite{Borges2010}. In conventional MRI, the brainstem nerve segment can only be deduced from its surrounding structures \cite{Casselman2008}. Sequences that can visualize myelinated structures are chosen, these include T2-weighted MRI  sequences such as FIESTA and CISS. Manual segmentation of T1 and T2 MRI requires great effort, with mixed results \cite{Miller2008a}. Diffusion MRI provided fibre orientation, and early attempts to segment the trigeminal system showed promise \cite{Upadhyay2008,Habas2007f}. Additionally, DTI showed strong ability to visualize spatial relationships in white matter anatomy, for example in delineating fornix sub-regions \cite{Chen2015c}, and in visualizing the distortion of CN VII/VIII by vestibular schwannomas \cite{Chen2011b}.
